Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-05-28 text/sections/01-introduction.tex
Finished adjusting the writing.  It is ready to Zenodo the data files.

Per student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per student in 1990 and \$8,300 in 2021, while the number of lecturers per student increased by 109\% and professors per student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported supplementary data on professor hiring at US public universities over 2010--2021.
% Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \citet[IPEDS]{ipeds} to study the impact of changes in state funding for all US public universities 1990--2021.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citet[IBHED]{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 per student (or a 10\% increase), do they change how many faculty members they employ?
If so, which positions do universities substitute away from, and do these funding cuts affect the faculty themselves?

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This happened in 2010 when the University of Illinois System furloughed university staff following budget cuts after the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition \citep{hoxby2009changing,urquiola2020markets}, may notice that their rankings are negatively affected by falling professor--lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guarantee a substitution towards lecturers at US public universities.

@@

In my university-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for up to ten years after the initial funding cut, showing that effects are persistent and not transitory.
Over the same time period, state funding per student fell by around 35\%, professors per student fell by 9\%, and lecturers per student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
This implies that faculty substitution arose by limiting the hiring of new professors, and is supported by suggestive evidence on professor hiring counts.

@@

This paper proceeds as follows.
\autoref{sec:data} describes the data for university funding, data on faculty in Illinois, and recent trends in higher education funding.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty, with \autoref{sec:results} presenting the empirical results.
%\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

>>>>

per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \citet[IPEDS]{ipeds} to study the impact of changes in state funding for all US public universities 1990--2021.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citet[IBHED]{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 per-student (or a 10\% increase), do they change how many faculty members they employ?
If so, which positions do universities substitute away from, and do these funding cuts affect the faculty themselves?

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This happened in 2010 when the University of Illinois System furloughed university staff following budget cuts after the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition \citep{hoxby2009changing,urquiola2020markets}, may notice that their rankings are negatively affected by falling professor-lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guarantee a substitution towards lecturers at US public universities.

@@

In my university-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of 10\% per-student leads to a fall of 1.4\% in the number of assistant professors per-student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per-student.
Local projection estimates show that these effects linger for up to ten years after the initial funding cut, showing that effects are persistent and not transitory.
Over the same time period, state funding per-student fell by around 35\%, professors per-student fell by 9\%, and lecturers per-student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per-student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per-student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
This implies that faculty substitution arose by limiting the hiring of new professors, and is supported by suggestive evidence on professor hiring counts.

@@

This paper proceeds as follows.
\autoref{sec:data} describes the data for university funding, data on faculty in Illinois, and recent trends in higher education funding.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty, with \autoref{sec:results} presenting the empirical results.
\autoref{sec:conclusion} concludes.

### 2024-05-28 text/sections/02-data.tex
Finished adjusting the writing.  It is ready to Zenodo the data files.

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process \citep{NBERw23736}.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state funding, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (e.g., caused by changing economic conditions) necessarily affect state expenditures.
Public universities have lower lobbying power than other state institutions, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than that of 1970 or 1972, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
These features lead to yearly variation in state funding not seen in other revenues sources, such as federal funding.

State funding for public universities stagnated over 1990--2021.
\autoref{fig:funding} show the trends in revenues for the mean public university for the years 1990--2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state funding stagnated at around \$100 million per year-university for 1990--2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with an average of 40 students for every one full professor, and little change thereafter.
Public universities start with 40 students per professor, and by 2021 there are 52.4 students professors for every full professor --- with the largest rise coming in the 2008--2011 time period.
The general trend is similar for the number of assistant professors at both public and private universities.
Private and public universities have similar numbers of associate and full professors before the year 2000, yet this number fell by over 20\% afterwards only for public universities: in 2021 the average public university has 6 fewer full professors per hundred students than the average private university.
Over the same time period, we see the rise in use of lecturers; lecturers were were employed at similar rates in private and public sectors in 1990, but have public universities increased their reliance on lecturers more than private universities ever since.

\subsection{Trends in Illinois}
\label{sec:trends-illinois}
Illinois funding for higher education has stagnated, and has experienced serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding Illinois public universities fell by over 50\% over 2010--2021, in both absolute and per-student terms (see \autoref{fig:illinois-funding}).
%Similar to the figures for the rest of the country, there has been a corresponding substitution towards tuition revenue.

% Paragraph describing how IL is similar to rest of country.

There was not only a stagnation in state funding in this time period, but also large annual rises and falls, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While most public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This means that Illinois public universities exhibit sizable changes in their state funding over 2010--2021, of similar order to those for the rest of the country over 1990--2021.
Additionally, the 2016 episode stemmed entirely from political disagreements, and not from state decisions regarding higher education and its finances \citep{young2020squandered}, exhibiting how state-level changes in funding affect public universities thanks to unrelated issues.

>>>>

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process \citep{NBERw23736}.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state funding, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obliged to run a balanced-budget, so that yearly variation in tax revenues (e.g., caused by changing economic conditions) necessarily led to state government spending cuts.
Public universities have lower lobbying power than other state institutions, so often bear the brunt of these funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than that of 1970 or 1972, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
These features lead to yearly variation in state funding not seen in other revenues sources, such as federal funding.

State funding for public universities stagnated over the last thirty years.
\autoref{fig:funding} show the trends in revenues for the mean public university for the years 1990--2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state funding stagnated at around \$100 million per year-university for 1990--2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with an average of 40 students for every one full professor, and little change thereafter.
Public universities start with 40 students per professor, and by 2021 there are 52.4 students professors for every full professor --- with the largest rise coming in the 2008--2011 time period.
The general trend is similar for the number of assistant professors at both public and private universities.
Private and public universities have similar numbers of associate and full professors before the year 2000.
After 2000, the number of professors at public univeristies fell drastically, with the average public university has 6 fewer full professors per hundred students than the average private university by 2021.
Over the same time period, we see the rise in use of lecturers; lecturers were employed at similar rates in private and public sectors in 1990.
Since 1990, both sectors began employing more lecturers on a per-student basis.
In sum, this means public universities have begun relying on lecturers, because they decreased their professor counts, compared to private unviersities who started employing more lecturers and professors.

\subsection{Trends in Illinois}
\label{sec:trends-illinois}
Illinois funding for higher education has stagnated, experiencing serious declines in the decade 2010--2021 --- similar in magnitude to the nation-wide decline over 1990--2021.
State funding Illinois public universities fell by over 50\% over 2010--2021, in both absolute and per-student terms (see \autoref{fig:illinois-funding}).

There was not only a stagnation in state funding in this time period, but also large annual rises and falls, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While most public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This means that Illinois public universities exhibit sizable changes in their state funding over 2010--2021, of similar order to those for the rest of the country over 1990--2021.
Additionally, the 2016 episode stemmed entirely from political disagreements, and not from state decisions regarding higher education and its finances, exhibiting how state-level changes in funding affect public universities thanks to unrelated issues \citep{young2020squandered}.

### 2024-05-28 text/sections/03-concepts.tex
Finished adjusting the writing.  It is ready to Zenodo the data files.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\section{Conceptual Framework}
\label{sec:conceptual}

Universities can respond multiple ways to state funding cuts.
\cite{NBERw23736} established that public universities experiencing state funding cuts did not fully off-set by raising tuition, and instead cut spending.
It is not clear how these funding or spending cuts went on to affect faculty --- or whether they affected faculty at all.

Wider changes in the American higher education sector could concurrently explain the trends in state funding and faculty substitution.
College selectivity changed drastically over the 21st century, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, leaving public universities substituting toward lecturers over time.
This is one way in which the trends may be concurrent, but not causally related.
I address this issue by using shift-share IV approach to isolate funding cuts that affected more funding-intensive universities, measuring the rate of substitution between lecturers and professors in response to yearly changes in state-wide funding.

Universities could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020 at the onset of the Covid-19 pandemic.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the salaries public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting offers at public universities.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time --- see \cite{blackaby2005} for an overview of faculty outside options.
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: lecturers have substantially lower annual salaries than professors.
The average Illinois assistant professor's annual salary is $\$76,897$, tenured professors' is $\$109,283$, while lecturers' is $\$31,449$ (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
If a public university's primary obligation is to teach, and they must fulfil this objective with less funding, then substituting away from tenure-track and tenured professors towards lecturers is rationalisable purely on this basis.

If faculty are affected by state funding cuts, then it is still not clear whether effects are persistent or merely transitory.
Universities employ professors on multiple-year contracts (e.g., the tenure contract), and professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for effects to trickle down to faculty.
On the other hand, university departments hire in annual cycles, so that a funding cut in one year may immediately impact faculty hiring in the same year, and thus faculty counts in the next academic year.
% Yet, it could be the case that faculty hiring resumes the very next year, or even hires more to compensate for last year's cancelled hiring cycle.
%This means that state funding cuts could effect faculty in only transitory changes in faculty counts.
As such, this paper investigates the dynamics of state funding cuts and their effects on faculty, by employing local projections methods to measure the persistence of effects for multiple years following a state funding cut.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\section{Conceptual Framework}
\label{sec:conceptual}
University administrations can respond in multiple ways to state funding cuts.
\cite{NBERw23736} established that public universities experiencing state funding cuts did not fully off-set by raising tuition, and instead cut spending.
It is not clear how these funding or spending cuts went on to affect faculty --- or whether they affected faculty at all.

Wider changes in American higher education could concurrently explain the national trends in state funding and faculty substitution.
College selectivity changed drastically over the 21st century, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, leaving public universities substituting toward lecturers over time.
This is one way in which the trends may be concurrent, but not causally related.
I address this issue by using shift-share IV approach to isolate funding cuts that affected more funding-intensive universities, measuring the rate of substitution between lecturers and professors in response to yearly changes in state-wide funding.

Similarly, institution-specific factors could explain correlations between state funding cuts and individual-faculty outcomes, such as salaries or rates of tenure.
FOr example, an administration could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020 at the onset of the Covid-19 pandemic.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.
}
Funding cuts could also limit the salaries public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting offers at public universities.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time --- see \cite{blackaby2005} for an overview of faculty outside options.
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: lecturers have substantially lower annual salaries than professors.
The average Illinois assistant professor's annual salary is $\$76,897$, tenured professors' is $\$109,283$, while lecturers' is $\$31,449$ (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
If a public university's primary obligation is to teach, and they must fulfil this objective with less funding, then substituting away from tenure-track and tenured professors towards lecturers is rationalisable purely on this basis.

If faculty are affected by state funding cuts, then it is still not clear whether effects are persistent or merely transitory.
Universities employ professors on multiple-year contracts (e.g., the tenure contract), and professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for effects to trickle down to faculty.
On the other hand, university departments hire in annual cycles, so that a funding cut in one year may immediately impact faculty hiring in the same year, and thus faculty counts in the next academic year.
% Yet, it could be the case that faculty hiring resumes the very next year, or even hires more to compensate for last year's cancelled hiring cycle.
%This means that state funding cuts could effect faculty in only transitory changes in faculty counts.
As such, this paper investigates the dynamics of state funding cuts, and their effects on faculty, by employing local projections methods to measure the persistence of effects for multiple years after a state funding cut.
